
```{r setup, include=FALSE}
library(valleybikeData)
library(tidyverse)
library(curl)
```

# Demographic Exploration of ValleyBike Users

## Motivations and initial summary

The motivation behind this project is that it has only recently become possible to obtain demographic user data and connect it to the existing data on the number of trips. Thus, it has become possible to gather information on whether different demographic factors influence frequency and duration of trips. This can help ValleyBike leadership understand who exactly are its largest clients and what they use the bikes for, which can then influence the location of future stations and other changes to the ValleyBike structure.

In total, there are 16589 users and 112011 trips by those users from June 2018 to October 2020. Trips shorter than 2 minutes or longer than 4 hours were excluded from this analysis. The central question of this project is what patterns we can glean from different demographic information about users, as well as some of the trip characteristics that haven't been explore before, e.g. # of trips by hour and whether the day of the trip falls on a weekend/holiday.

## Trips to NA General Summary

```{r, warning = FALSE, echo = FALSE, message = FALSE}
trips_to_NA <- trips %>% filter(is.na(end_station) & !is.na(start_station))
cutoff <- 5000
cleaned_trips_NA2 <- trips_to_NA %>% filter(duration < cutoff, duration > 120)

NA_user <- trips_to_NA %>%
  group_by(user_id) %>%
  tally(sort = TRUE)
list_non_super_NA_users <- filter(NA_user, n < 20)$user_id
cleaned_trips_NA3 <- cleaned_trips_NA2 %>% filter(user_id %in% list_non_super_NA_users)

trips_per_day <- trips %>% 
  mutate(start_date = cut.POSIXt(start_time,"day")) %>% 
  count(start_date) 

cleaned_trips_NA3_count <- cleaned_trips_NA3 %>% 
  mutate(start_date = cut.POSIXt(start_time,"day")) %>% 
  count(start_date) %>%
  rename( na_per_day = n) %>%
  left_join(trips_per_day) %>%
  mutate(start_date = as.POSIXct(start_date,"")) %>%
  filter(na_per_day/n < 0.2)
```

```{r, warning = FALSE}
ggplot(data = cleaned_trips_NA3_count) +
  geom_bar(aes(x = start_date, y = na_per_day/n),stat="identity") +
  labs(title = "Barplot of Normalized Rides to NA Chronologically") +
  xlab("Ride Start Date") + ylab("Proportion of all rides that are from station to NA")

```

```{r, echo = FALSE, message=FALSE}
users2020 <- readr::read_csv(curl("https://raw.githubusercontent.com/DSC-WAV/proj-valleyriders/master/data/users_2020_10.csv?token=ANCR62LHPFUAQPVKBTFBOFK73JU22")) %>% 
  mutate(
    membership_start = lubridate::parse_date_time(membership_start, "YmdHMS"),
    membership_end = lubridate::parse_date_time(membership_end, "YmdHMS")) %>%
  select(-X1)

get_other <- function(ds) {
  categories <- c("Missing", "Agawam", "Belchertown", "Boston", "Chicopee", "East Longmeadow", "Enfield, CT",
                  "Granby", "Hadley", "Leverett", "Ludlow", "New York City", "South Hadley", "Southampton",
                  "Sunderland", "West Springfield", "Westfield")

  for (i in 1:length(ds$location)) {
    if (str_detect(ds$location[i], "Other") | ds$location[i] %in% categories) {
      ds$location[i] <- "Other"
    } 
  }
  return(ds)
}
users2020_cleaned <- get_other(users2020)

users2020_cleaned <- users2020_cleaned 

users2020_cleaned <- users2020_cleaned %>%
  mutate(age_group = ifelse(age_group == "<=23", "18-23", age_group),
         gender = case_when(
           is.na(gender) | gender == "o" ~ "Missing",
           gender == "f" ~ "Female",
           gender == "m" ~ "Male"))

max_duration <- 4 * 60 * 60    # max valid duration (in seconds)
min_duration <- 2 * 60     # minimum valid duration (in seconds)
trips_cleaned <- trips %>% 
  filter(duration >= min_duration, duration <= max_duration)

trips_combined <- left_join(trips_cleaned, users2020_cleaned, by = "user_id")

trips_cleaned <- trips_combined %>% 
  filter(!is.na(location)) #, end_time < as.POSIXct("2020-01-01")) #Important

#Time zone adjustment?
holidates <- lubridate::ymd(c("2018-07-04", "2019-07-04", "2020-07-04", 
                                  "2018-09-03","2019-09-02", "2020-09-07", 
                                  "2018-05-28", "2019-05-27", "2020-05-25"))
trips_cleaned <- trips_cleaned %>%
  mutate(type_of_day = ifelse(lubridate::wday(start_time, 
              label = TRUE) == "Sat" | lubridate::wday(start_time, 
              label = TRUE) == "Sun" | cut.POSIXt(start_time, 
                  "day") %in% holidates, "Weekend/Holiday", "Workday"))
  
      
trips_cleaned_users <- trips_cleaned %>%
  mutate(duration = duration/60, 
         age_group = ifelse(age_group == "<=23", "18-23", age_group),
         gender = case_when(
           is.na(gender) | gender == "o" ~ "Missing",
           gender == "f" ~ "Female",
           gender == "m" ~ "Male"))

#Potentially clean up membership_name variable
#unique(trips_cleaned$membership_name)

duration_by_user <- trips_cleaned_users %>%
  group_by(user_id) %>%
  summarize(med_duration = median(duration))

#delete line just below?
summary_by_user <- left_join(duration_by_user, users2020_cleaned, by = "user_id")

trips_cleaned_users <- trips_cleaned_users %>% 
  mutate(Hour = lubridate::hour(start_time))
```


\newpage
## Tables and Plots of Trips by User Age

```{r, warning = FALSE, message=FALSE, fig.height= 6, fig.width = 8, echo = FALSE}
age_table <- users2020_cleaned %>% summarise(age_group) %>% count(age_group) %>%
  rename('Age Group' = age_group, 'Count_user' = n) %>%
  mutate(`Proportion of Users` = round(Count_user/sum(Count_user), 2)) %>%
  janitor::adorn_totals("row")

tempo <- trips_cleaned_users %>% group_by(user_id, age_group) %>% summarise(trip_number = n())

median_ages <- tempo %>% 
  group_by(age_group) %>% 
  summarize(`Median Number of Trips Per User` = median(trip_number)) %>%
  rename('Age Group' = age_group)

trip_age_table <- trips_cleaned_users %>% 
  summarise(age_group, `Median Duration` = median(duration)) %>% 
  count(age_group) %>%
  rename('Age Group' = age_group, 'Count_trip' = n) %>%
  mutate(`Proportion of Trips` = round(Count_trip/sum(Count_trip), 2)) %>%
  janitor::adorn_totals("row")

duration_age_table <- trips_cleaned_users %>% 
  group_by(age_group) %>% 
  summarise(`Median Trip Duration, mins` = round(median(duration),2)) %>%
  rename('Age Group' = age_group)


age_table_comb <- full_join(age_table, trip_age_table, by = "Age Group") %>% 
  full_join(median_ages, by = "Age Group") %>%
  mutate(`Avg Number of Trips per User`  = round(Count_trip/Count_user,2)) %>%
  select(-Count_trip, -Count_user) %>% 
  full_join(duration_age_table, by = "Age Group") 

age_table_comb[7, "Proportion of Trips"] <- 1
age_table_comb[7, "Median Trip Duration, mins"] <- median(round(trips_cleaned_users$duration, 2))
age_table_comb[7, "Median Number of Trips Per User"] <- median(tempo$trip_number)

knitr::kable(age_table_comb %>% arrange(`Age Group`), 
             caption = "Trip Summary By User Age")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour), bins = 24,colour='black', fill='gray') + 
  facet_wrap(~age_group) +
  ylab("Number of Trips by Hour") +
  ggtitle("Histogram of Trips per Hour by Age Group")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour, y =
                       (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]), 
                 bins = 24,colour='black', fill='gray') + 
  facet_wrap(~age_group) +
  ylab("Number of Trips per Hour Normalized by Age Group") +
  ggtitle("Histogram of Trips per Hour Normalized by Age Group")
```

\newpage

## Tables and Plots by Gender

```{r, warning= FALSE, fig.height= 6, fig.width = 8, message=FALSE, echo = FALSE}

gender_table <- users2020_cleaned %>% 
  summarise(gender) %>% 
  count(gender) %>%
  rename(Gender = gender, 'Count_user' = n) %>%
  mutate(`Proportion of Users` = round(Count_user/sum(Count_user), 2)) %>%
  janitor::adorn_totals("row")

tempo <- trips_cleaned_users %>% group_by(user_id, gender) %>% summarise(trip_number = n())

median_ages <- tempo %>% 
  group_by(gender) %>% 
  summarize(`Median Number of Trips Per User` = median(trip_number)) %>%
  rename(Gender = gender)

trip_gender_table <- trips_cleaned_users %>% 
  summarise(gender) %>% 
  count(gender) %>%
  rename(Gender = gender, 'Count_trip' = n) %>%
  mutate(`Proportion of Trips` = round(Count_trip/sum(Count_trip), 2)) %>%
  janitor::adorn_totals("row")

duration_gender_table <- trips_cleaned_users %>% 
  group_by(gender) %>% 
  summarise(`Median Trip Duration, mins` = round(median(duration),2)) %>%
  arrange(`Median Trip Duration, mins`) %>%
  rename(Gender = gender)

gender_table_comb <- full_join(gender_table, trip_gender_table, 
                                 by = "Gender") %>%
  full_join(median_ages, by = "Gender") %>%
  mutate(`Avg Number of Trips per User`  = round(Count_trip/Count_user,2)) %>%
  select(-Count_trip, -Count_user) 

gender_table_comb <- full_join(duration_gender_table, 
                                 gender_table_comb, by = "Gender")
gender_table_comb[4, 
            "Median Trip Duration, mins"] <- 
median(round(trips_cleaned_users$duration, 2))

gender_table_comb[4, "Median Number of Trips Per User"] <- median(tempo$trip_number)

knitr::kable(gender_table_comb[, c(1,3,4,5,6, 2)], 
             caption = "Trip Summary By User Gender")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour), bins = 24,colour='black', fill='gray') + 
  facet_grid(rows = vars(gender)) +
  ylab("Number of Trips by User Gender") +
  ggtitle("Histogram of Trips per Hour by Gender")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour, y = 
            (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]), 
  bins = 24, colour='black', fill='gray') + 
  facet_grid(rows = vars(gender)) +
  ylab("Number of Trips by Gender of User (Normalized by Location)") +
  ggtitle("Histogram of Trips per Hour by Gender (Normalized by Location)")

```

It is important to note that, although 73% of ValleyBike registered users have not stated their gender, only 45% of all trips were taken by these users. Known female and male users, on the other hand, are only 7% and 20% of all users, but complete 13% and 42% of all trips, respectively.

\newpage

## Tables by Gender and Age

```{r, warning= FALSE,  message=FALSE, echo = FALSE}

gender_age_table <- table(summary_by_user$age_group, summary_by_user$gender)
knitr::kable(gender_age_table, caption = "Summary Table of Counts by Gender and Age")

dem_table <- users2020_cleaned %>% mutate(dem_type = paste(gender,age_group,sep=" ") ) %>%
  summarise(dem_type) %>% count(dem_type) %>%
  rename('Demographic Group' = dem_type, 'Count_user' = n) %>%
  mutate(`Proportion of Users` = round(Count_user/sum(Count_user), 3)) %>%
  janitor::adorn_totals("row")

tempo_dem <- trips_cleaned_users %>% 
  mutate(dem_type = paste(gender,age_group,sep=" ")) %>%
  group_by(user_id, dem_type) %>% 
  summarise(trip_number = n())

median_dem <- tempo_dem %>% 
  group_by(dem_type) %>% 
  summarize(`Median Number of Trips Per User` = median(trip_number)) %>%
  rename('Demographic Group' = dem_type)

trip_dem_table <- trips_cleaned_users %>% 
  mutate(dem_type = paste(gender,age_group,sep=" ")) %>%
  summarise(dem_type, `Median Duration` = median(duration)) %>% 
  count(dem_type) %>%
  rename('Demographic Group' = dem_type, 'Count_trip' = n) %>%
  mutate(`Proportion of Trips` = round(Count_trip/sum(Count_trip), 2)) %>%
  janitor::adorn_totals("row")

duration_dem_table <- trips_cleaned_users %>% 
  mutate(dem_type = paste(gender,age_group,sep=" ")) %>%
  group_by(dem_type) %>% 
  summarise(`Median Trip Duration, mins` = round(median(duration),2)) %>%
  rename('Demographic Group' = dem_type)


dem_table_comb <- full_join(dem_table, trip_dem_table, by = "Demographic Group") %>% 
  full_join(median_dem, by = "Demographic Group") %>%
  mutate(`Avg Number of Trips per User`  = round(Count_trip/Count_user,2)) %>%
  select(-Count_trip, -Count_user) %>% 
  full_join(duration_dem_table, by = "Demographic Group") 

dem_table_comb[19, "Proportion of Users"] <- 1
dem_table_comb[19, "Proportion of Trips"] <- 1
dem_table_comb[19, "Median Trip Duration, mins"] <- median(round(trips_cleaned_users$duration, 2))
dem_table_comb[19, "Median Number of Trips Per User"] <- median(tempo_dem$trip_number)

knitr::kable(dem_table_comb %>% arrange(`Demographic Group`), 
             caption = "Trip Summary By User Age and Gender")
```

If we look at both gender and age, we see that the patterns of the number of trips per user and median trip durations differ by gender. THe oldest user group (60+) takes the fewest trips for all three categories. However, for users whose gender is missing, the youngest users take the greatest number of trips. For males, users 30-40 years old take even more trips than users 18-23, but the youngest female users take much fewer trips on average, only slightly more than the oldest female users. However, since the female age group is the smallest, their summary numbers are more likely to fluctuate than the numbers for the larger male and missing groups. For trip durations, the oldest users take either average length trips (for males) or the longest trips (for females and especially for the missing group). Although the youngest users in male and missing categories take the shortest trips, the female bikers are on the high end, with median trip duration of 19.5 minutes, longer than 18-23 users in the other two groups and the second longest trip duration among female age groups. However, female users 23-30 have the shortest median trip duration out of all groups, which implies that the effect is caused by in-group vartiability and not by gender-specific biking trends. Since a significant number of trips involves secondary stops, the median duration would be higher if round trips with a stop get counted as one trip, while the average number of trips would be correspondingly lower.

## Tables and Plots by User Location

```{r, warning= FALSE, fig.height= 6, fig.width = 8, message=FALSE, echo = FALSE}
location_table <- users2020_cleaned %>% 
  summarise(location) %>% 
  count(location) %>%
  rename(Location = location, 'Count_user' = n) %>%
  mutate(`Proportion of Users` = round(Count_user/sum(Count_user), 2)) %>%
  janitor::adorn_totals("row")

tempo <- trips_cleaned_users %>% group_by(user_id, location) %>% summarise(trip_number = n())

median_ages <- tempo %>% 
  group_by(location) %>% 
  summarize(`Median Number of Trips Per User` = median(trip_number)) %>%
  rename('Location' = location)


trips_location_table <- trips_cleaned_users %>% 
  summarise(location) %>% 
  count(location) %>%
  rename(Location = location, 'Count_trip' = n) %>%
  mutate(`Proportion of Trips` = round(Count_trip/sum(Count_trip), 2)) %>%
  janitor::adorn_totals("row")

duration_location_table <- trips_cleaned_users %>% 
  group_by(location) %>% 
  summarise(`Median Trip Duration, mins` = round(median(duration),2)) %>%
  arrange(`Median Trip Duration, mins`) %>%
  rename(Location = location)


location_table_comb <- full_join(location_table, trips_location_table, 
                                 by = "Location") %>%
  full_join(median_ages, by = "Location") %>%
  mutate(`Avg Number of Trips per User`  = round(Count_trip/Count_user,2)) %>%
  select(-Count_trip, -Count_user) 

location_table_comb <- full_join(duration_location_table, 
                                 location_table_comb, by = "Location")
location_table_comb[7, 
            "Median Trip Duration, mins"] <- 
  median(round(trips_cleaned_users$duration, 2))

location_table_comb[7, "Median Number of Trips Per User"] <- median(tempo$trip_number)

knitr::kable(location_table_comb[, c(1,3,4,5,6, 2)], 
             caption = "Trip Summary By User Location")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour), bins = 24,colour='black', fill='gray') + 
  facet_wrap(~location) +
  ylab("Number of Trips by User Location") +
  ggtitle("Histogram of Trips per Hour by User Location")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour, y =
                       (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]), 
                 bins = 24,colour='black', fill='gray') + 
  facet_wrap(~location) +
  ylab("Number of Trips by Location (Normalized by Location)") +
  ggtitle("Histogram of Trips per Hour by User Location (Normalized by Location)")
```

\newpage

## Tables and Plots by User Type

```{r, warning= FALSE, fig.height= 6, fig.width = 8, message=FALSE, echo = FALSE}
type_table <- users2020_cleaned %>% 
  summarise(dtype) %>% 
  count(dtype) %>%
  rename('User Account Type' = dtype, 'Count_user' = n) %>%
  mutate(`Proportion of Users` = round(Count_user/sum(Count_user), 2)) %>%
  janitor::adorn_totals("row")

tempo <- trips_cleaned_users %>% group_by(user_id, dtype) %>% summarise(trip_number = n())

median_ages <- tempo %>% 
  group_by(dtype) %>% 
  summarize(`Median Number of Trips Per User` = median(trip_number)) %>%
  rename('User Account Type' = dtype)

trip_type_table <- trips_cleaned_users%>% 
  summarise(dtype) %>% 
  count(dtype) %>%
  rename('User Account Type' = dtype, 'Count_trip' = n) %>%
  mutate(`Proportion of Trips` = round(Count_trip/sum(Count_trip), 2)) %>%
  janitor::adorn_totals("row")

duration_type_table <- trips_cleaned_users %>% 
  group_by(dtype) %>% 
  summarise(`Median Trip Duration, mins` = round(median(duration),2)) %>%
  arrange(`Median Trip Duration, mins`) %>%
  rename('User Account Type' = dtype)

type_table_comb <- full_join(type_table, trip_type_table, 
                                 by = "User Account Type") %>%
  full_join(median_ages, by = "User Account Type") %>%
  mutate(`Avg Number of Trips per User`  = round(Count_trip/Count_user,2)) %>%
  select(-Count_trip, -Count_user) 

type_table_comb <- full_join(duration_type_table, 
                                 type_table_comb, by = "User Account Type")
type_table_comb[4, 
            "Median Trip Duration, mins"] <- 
median(round(trips_cleaned_users$duration, 2))
type_table_comb[4, "Median Number of Trips Per User"] <- median(tempo$trip_number)

knitr::kable(type_table_comb[, c(1,3,4,5,6, 2)], 
             caption = "Trip Summary By User Account Type")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour), bins = 24,colour='black', fill='gray') + 
  facet_grid(rows = vars(dtype)) +
  ylab("Number of Trips by User Account Type") +
  ggtitle("Histogram of Trips per Hour by User Account Type")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour, y =
                       (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]), 
bins = 24, colour='black', fill='gray') + 
  facet_grid(rows = vars(dtype)) +
  ylab("Proportion of Trips by User Account Type (Normalized per Account Type)") +
  ggtitle("Histogram of Trips per Hour Normalized by User Account Type")
```

\newpage
## Tables and Plots by Whether the Trip is on a Workday

```{r, warning= FALSE, fig.height= 6, fig.width = 8, message=FALSE, echo = FALSE}
day_table <- trips_cleaned_users%>% 
  summarise(type_of_day) %>% 
  count(type_of_day) %>%
  rename('Type of Day' = type_of_day, 'Count' = n) %>%
  mutate(`Proportion of Trips` = round(Count/sum(Count), 2)) %>%
  janitor::adorn_totals("row")

duration_day_table <- trips_cleaned_users %>% 
  group_by(type_of_day) %>% 
  summarise(`Median Trip Duration, mins` = round(median(duration),2)) %>%
  arrange(`Median Trip Duration, mins`) %>%
  rename('Type of Day' = type_of_day)

day_table_comb <- full_join(duration_day_table,
                            day_table, by = "Type of Day") %>% 
  select(-Count)

day_table_comb[3, 
            "Median Trip Duration, mins"] <- 
median(round(trips_cleaned_users$duration, 2))

knitr::kable(day_table_comb[, c(1,3,2)], 
             caption = "Trip Summary By Type of Day")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour), bins = 24,colour='black', fill='gray') + 
  facet_grid( rows = vars(type_of_day)) +
  ylab("Number of Trips by Whether Trip is on Weekend") +
  ggtitle("Histogram of Trips per Hour by Whether The Trip is on a Workday")

ggplot(trips_cleaned_users) +
  geom_histogram(aes(x = Hour, y = 
                       (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]), 
                 bins = 24,colour='black', fill='gray') + 
  facet_grid( rows = vars(type_of_day)) +
  ylab("Number of Trips by Whether Trip is on a Workday") +
  ggtitle("Histogram of Trips per Hour by Whether The Trip is on a Workday")
```

Note: other than weekends, the holidays include Independence Day, Labor Day, and Memorial Day.

\newpage
## Table Summary of Most Common Visited Stations by User Location

Tables included below list the number of visited stations (as either a starting station or an end station). The proportion of trips variable is the proportion of all trips for users who put that place as their location which includes this station. Since trips are between two, usually different, locations, the total of the proportion of all trips may add up to > 1.

```{r, results='asis', warning = FALSE, message=FALSE, echo = FALSE}
trips_cleaned_p1 <- trips_cleaned %>% 
  select(-end_station) %>%
  rename(station = "start_station")
trips_cleaned_p2 <- trips_cleaned %>% 
  select(-start_station) %>%
  rename(station = "end_station")

trips_cleaned_stations <- rbind(trips_cleaned_p1, trips_cleaned_p2)

locations <-  unique(trips_cleaned_stations$location)
station_cities <- stations %>% select(name, city)
 
for (i in 1:(length(locations)-1)) {
  temp <- trips_cleaned_stations %>% filter(location == locations[i])
  total_n_trips <- nrow(trips_cleaned %>% 
                          filter(location == locations[i]))
  temp_city <- temp %>% 
  group_by(station) %>%
  tally(sort = TRUE) %>% 
    top_n(20, wt = n) %>%
    left_join(station_cities, by = c("station" = "name")) %>%
    rename(`total number of trips` = n) %>% 
    mutate(`proportion of trips` = round(`total number of trips` / total_n_trips,3))
  
  temp2 <- knitr::kable(temp_city[c(1,3,2, 4)], 
                        caption = paste("User Location: ", locations[i], sep = ""))
  cat("\n\n\\pagebreak\n")
  print(temp2)
}

```

\newpage

## Appendix

```{r}
table(summary_by_user$dtype, summary_by_user$membership_type)
table(summary_by_user$dtype, summary_by_user$gender)
```

```{r, message = FALSE, warning = FALSE}
knitr::kable(table(summary_by_user$membership_name, summary_by_user$dtype)) 
```

```{r, message = FALSE, warning = FALSE}
knitr::kable(table(summary_by_user$membership_name, summary_by_user$membership_type)) %>% 
  kableExtra::column_spec(1, "2in")

table(summary_by_user$membership_name, summary_by_user$gender)

table(summary_by_user$location, summary_by_user$dtype)
table(summary_by_user$location, summary_by_user$membership_type)
gender_loc_table <- table(summary_by_user$location, summary_by_user$gender)
gender_loc_table
ftable(round(prop.table(gender_loc_table, margin = 1),3))
table(summary_by_user$age_group, summary_by_user$dtype)
```

